\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-05-31 paper/ms.tex
fixing todos macros

In this paper, we present a method for computing a class of GP models that
scales linearly with the number of data points $\mathcal{O}(N)$ for one
dimensional data sets.
\todo{This method is a generalization of a method developed by
\citet{Ambikasaran:2015}
 ({\bf This is no longer correct -different Ambikasaran result!})}
that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
This method can only be used with one-dimensional datasets and the covariance
function must be represented by a mixture of exponentials; we will return to a
discussion of what this means in detail in the following sections.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic (non-uniform).

This method achieves linear scaling by exploiting structure in the covariance
matrix when it is generated by a mixture of exponentials.
\todo{More precisely, the matrix can be embedded in an equivalent higher-dimensional
system that can be solved efficiently using a sparse or banded solver. Replace
with(?): The matrix can be expressed as the sum of semi-separable matrices, each
of which is constructed from the outer product of one-dimensional vectors.  These
are a type of low-rank matrix, which may be solved by Cholesky decomposition
using a novel technique developed by one of us (SA).}
As we will discuss in the following pages, this choice of kernel function is
not as restrictive as it first seems because it arises naturally in some
physical systems and it can be used as an effective model in many other cases.
This method is especially appealing compared to other similar methods~--~we
return to these below~--~because it is exact, flexible, simple, and fast.

@@

In the following pages, we motivate the general problem of GP modeling,
describe the \todo{(delete) previously published} scalable method
\citep{Rybicki:1995, Ambikasaran:2015} and our generalization, and demonstrate
the model's application on various real and simulated data sets.
Alongside this paper, we have released well-tested and documented open source
implementations written in \project{C++}, \project{Python}, and
\project{Julia}.
These are available online at \project{GitHub}
\url{https://github.com/dfm/celerite} and \project{Zenodo}
\citep{Foreman-Mackey:2017}.

@@

GPs are stochastic models consisting of a mean function
$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
function $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ parameterized by
$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, \response{the log-likelihood function $\mathcal{L}
(\bvec{\theta},\,\bvec{\alpha})$} with a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln \mathcal{L} (\bvec{\theta},\,\bvec{\alpha}) =
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
\eq{gp-likelihood} is the equation of an $N$-dimensional Gaussian and it can
be derived as the generalization of what we astronomers call ``$\chi^2$''
to include the effects of correlated noise.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ \todo{(Why
isn't X boldface?)} can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.

@@

To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
compute the log-determinant of $K$ and solve linear systems of the form
$K\,\bvec{b}=\bvec{y}$ for $\bvec{b}$, where $K$ is a rank-$2\,J$
semi-separable matrix.
If $K$ is positive semi-definite, these can be computed using the Cholesky
factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,X^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $X$ is an unknown
$N \times 2\,J$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,X^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,X^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,X^\T)\,D
    + D\,\mathrm{triu}(X\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,X^\T)\,D\,\mathrm{triu}(X\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $X$. \todo{X was used to represent independent
parameter matrix in definition of likelihood.}
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,X_{n-1,j}\,X_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
X_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
$S_{1,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
observation~--~so the total cost of this factorization scales as
$\mathcal{O}(N\,J^2)$.

@@

The data are simulated from a quasiperiodic GP with the kernel
\begin{eqnarray}\eqlabel{sim-wrong-true}
k_\mathrm{true} (\tau) = \alpha\,
    \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{true}}\right)
\end{eqnarray}
where $P_\mathrm{true}$ is the fundamental period of the process.
This autocorrelation structure corresponds to the power spectrum
\citep{Wilson:2013}
\begin{eqnarray}
S_\mathrm{true} (\omega) = \frac{\lambda\,\alpha}{2}\,\left[
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right) +
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega+
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right)
\right]
\end{eqnarray}
which, for large values of $\omega$, falls off exponentially. \todo{$\alpha$
means the parameter vector in definition of likelihood.}
When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
$\omega^{-4}$ at most~--~it is clear that a \celeriteterm\ model can never
exactly reproduce the structure of this process.
That being said, we demonstrate that robust inferences can be made about
$P_\mathrm{true}$ even with this effective model.

@@

A source of variability that can be measured from time series measurements of
stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
\end{eqnarray}
where $P$ is the rotation period of the star.
GP modeling with the same kernel function has been proposed as a method of
measuring the mean periodicity in quasiperiodic photometric time series in
general \citep{Wang:2012}.
The key difference between \eq{sine2} and other quasiperiodic kernels is that
it is positive for all values of $\tau$.
We can construct a simple \celeriteterm\ covariance function with similar properties
as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{a}{2+b}\,e^{-\tau/c}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
\right]
\end{eqnarray}
for $a>0$, $b>0$, and $c>0$. \todo{a,b,c already have different meaning
in celerite kernel.}
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a substantial gain in computational efficiency.

@@

To demonstrate the method, we use a simple heuristic model where the PSD is
given by a mixture of 8 components with amplitudes and frequencies specified
by $\nu_\mathrm{max}$, $\Delta \nu$, and several nuisance parameters.
The first term is used to capture the granulation ``background''
\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
$S_g$ and $\omega_g$.
The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
parameter shared between terms and the frequencies are given by
\begin{eqnarray}
\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
\end{eqnarray}
and the amplitudes are given by
\begin{eqnarray}
S_{0,\,j} =
    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
\end{eqnarray}
where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
are shared nuisance parameters. \todo{A was used for diagonal of
covariance matrix.}
This model could be extended to include more angular modes and
$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by the fundamental
physical parameters of the star.

@@

\todo{Table 5 has two parameter with name "b":  the celerite coefficient,
and the impact parameter.}

@@

\todo{The D in Table 6 may be confused with the D used in the Cholesky solver.}

>>>>

In this paper, we present a method for computing a class of GP models that
scales linearly with the number of data points $\mathcal{O}(N)$ for one
dimensional data sets.
\dfmtodo{This method is a generalization of a method developed by
\citet{Ambikasaran:2015}
 ({\bf This is no longer correct -different Ambikasaran result!})}
that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
This method can only be used with one-dimensional datasets and the covariance
function must be represented by a mixture of exponentials; we will return to a
discussion of what this means in detail in the following sections.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic (non-uniform).

This method achieves linear scaling by exploiting structure in the covariance
matrix when it is generated by a mixture of exponentials.
\dfmtodo{More precisely, the matrix can be embedded in an equivalent higher-dimensional
system that can be solved efficiently using a sparse or banded solver. Replace
with(?): The matrix can be expressed as the sum of semi-separable matrices, each
of which is constructed from the outer product of one-dimensional vectors.  These
are a type of low-rank matrix, which may be solved by Cholesky decomposition
using a novel technique developed by one of us (SA).}
As we will discuss in the following pages, this choice of kernel function is
not as restrictive as it first seems because it arises naturally in some
physical systems and it can be used as an effective model in many other cases.
This method is especially appealing compared to other similar methods~--~we
return to these below~--~because it is exact, flexible, simple, and fast.

@@

In the following pages, we motivate the general problem of GP modeling,
describe the \dfmtodo{(delete) previously published} scalable method
\citep{Rybicki:1995, Ambikasaran:2015} and our generalization, and demonstrate
the model's application on various real and simulated data sets.
Alongside this paper, we have released well-tested and documented open source
implementations written in \project{C++}, \project{Python}, and
\project{Julia}.
These are available online at \project{GitHub}
\url{https://github.com/dfm/celerite} and \project{Zenodo}
\citep{Foreman-Mackey:2017}.

@@

GPs are stochastic models consisting of a mean function
$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
function $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ parameterized by
$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, \response{the log-likelihood function $\mathcal{L}
(\bvec{\theta},\,\bvec{\alpha})$} with a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln \mathcal{L} (\bvec{\theta},\,\bvec{\alpha}) =
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
\eq{gp-likelihood} is the equation of an $N$-dimensional Gaussian and it can
be derived as the generalization of what we astronomers call ``$\chi^2$''
to include the effects of correlated noise.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.

@@

To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
compute the log-determinant of $K$ and solve linear systems of the form
$K\,\bvec{b}=\bvec{y}$ for $\bvec{b}$, where $K$ is a rank-$2\,J$
semi-separable matrix.
If $K$ is positive semi-definite, these can be computed using the Cholesky
factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,X^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $X$ is an unknown
$N \times 2\,J$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,X^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,X^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,X^\T)\,D
    + D\,\mathrm{triu}(X\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,X^\T)\,D\,\mathrm{triu}(X\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $X$. \dfmtodo{X was used to represent independent
parameter matrix in definition of likelihood.}
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,X_{n-1,j}\,X_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
X_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
$S_{1,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
observation~--~so the total cost of this factorization scales as
$\mathcal{O}(N\,J^2)$.

@@

The data are simulated from a quasiperiodic GP with the kernel
\begin{eqnarray}\eqlabel{sim-wrong-true}
k_\mathrm{true} (\tau) = \alpha\,
    \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{true}}\right)
\end{eqnarray}
where $P_\mathrm{true}$ is the fundamental period of the process.
This autocorrelation structure corresponds to the power spectrum
\citep{Wilson:2013}
\begin{eqnarray}
S_\mathrm{true} (\omega) = \frac{\lambda\,\alpha}{2}\,\left[
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right) +
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega+
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right)
\right]
\end{eqnarray}
which, for large values of $\omega$, falls off exponentially. \dfmtodo{$\alpha$
means the parameter vector in definition of likelihood.}
When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
$\omega^{-4}$ at most~--~it is clear that a \celeriteterm\ model can never
exactly reproduce the structure of this process.
That being said, we demonstrate that robust inferences can be made about
$P_\mathrm{true}$ even with this effective model.

@@

A source of variability that can be measured from time series measurements of
stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
\end{eqnarray}
where $P$ is the rotation period of the star.
GP modeling with the same kernel function has been proposed as a method of
measuring the mean periodicity in quasiperiodic photometric time series in
general \citep{Wang:2012}.
The key difference between \eq{sine2} and other quasiperiodic kernels is that
it is positive for all values of $\tau$.
We can construct a simple \celeriteterm\ covariance function with similar properties
as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{a}{2+b}\,e^{-\tau/c}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
\right]
\end{eqnarray}
for $a>0$, $b>0$, and $c>0$. \dfmtodo{a,b,c already have different meaning
in celerite kernel.}
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a substantial gain in computational efficiency.

@@

To demonstrate the method, we use a simple heuristic model where the PSD is
given by a mixture of 8 components with amplitudes and frequencies specified
by $\nu_\mathrm{max}$, $\Delta \nu$, and several nuisance parameters.
The first term is used to capture the granulation ``background''
\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
$S_g$ and $\omega_g$.
The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
parameter shared between terms and the frequencies are given by
\begin{eqnarray}
\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
\end{eqnarray}
and the amplitudes are given by
\begin{eqnarray}
S_{0,\,j} =
    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
\end{eqnarray}
where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
are shared nuisance parameters. \dfmtodo{A was used for diagonal of
covariance matrix.}
This model could be extended to include more angular modes and
$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by the fundamental
physical parameters of the star.

@@

\dfmtodo{Table 5 has two parameter with name "b":  the celerite coefficient,
and the impact parameter.}

@@

\dfmtodo{The D in Table 6 may be confused with the D used in the Cholesky solver.}
